---
output: github_document
---

Tim Barry

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `sceptre` Nextflow pipeline

This repository contains the `sceptre` Nextflow pipeline. The `sceptre` Nextflow pipeline is a command line utility that facilitates running `sceptre` (i) out-of-core on a laptop or desktop or (ii) in a distributed fashion on a cluster or cloud. The `sceptre` Nextflow pipeline is highly scalable and memory-efficient; therefore, we recommend using the `sceptre` Nextflow pipeline instead of the R interface to `sceptre` in most cases.

# Requirements

-   Install [Nextflow](https://www.nextflow.io/)
-   Install the `sceptre` and `ondisc` packages:

<!-- -->

    # in R
    install.packages("devtools")
    devtools::install_github("katsevich-lab/sceptre")
    devtools::install_github("timothy-barry/ondisc")

-   Install the the `sceptre` nextflow pipeline.

<!-- -->

    # on command line 
    nextflow pull timothy-barry/sceptre-pipeline

# Pipeline arguments

We describe the `sceptre` pipeline arguments here.

## Input files

Four input files are required: (i) the multimodal ondisc matrix metadata file (`multimodal_metadata_fp`), (ii) the backing .odm file of the gene ondisc matrix (`gene_odm_fp`), (iii) the backing .odm file of the gRNA ondisc matrix (`gRNA_odm_fp`), and (iv) a data frame containing the set of gene-gRNA group pairs to analyze (`pair_fp`). On my (Tim's) machine these files are located in the following places:

```{r}
# in R
gasp_offsite_dir <- .get_config_path("LOCAL_GASPERINI_2019_DATA_DIR")
multimodal_metadata_fp <- paste0(gasp_offsite_dir, "at-scale/processed/multimodal/multimodal_metadata.rds")
gene_odm_fp <- paste0(gasp_offsite_dir, "at-scale/processed/gene/gasp_scale_gene_expressions.odm")
gRNA_odm_fp <- paste0(gasp_offsite_dir, "at-scale/processed/gRNA_ungrouped/gasp_scale_gRNA_counts_ungrouped.odm")
pair_fp <- paste0(gasp_offsite_dir, "at-scale/processed/multimodal/pairs.rds")
```

The multimodal ondisc matrix should satisfy the following conditions.

-   The multimodal ondisc matrix should have modalities named "gene" and "gRNA".

```{r}
# in R
library(ondisc)
crispr_experiment <- read_multimodal_odm(odm_fps = c(gene_odm_fp, gRNA_odm_fp),
                                         multimodal_metadata_fp = multimodal_metadata_fp)
gene_modality <- get_modality(crispr_experiment, "gene")
gRNA_modality <- get_modality(crispr_experiment, "gRNA")
```

-   The gRNA modality should be an integer-valued matrix of gRNA expressions or (less commonly) a logical matrix of gRNA-to-cell assignments. The feature covariate matrix of the gRNA modality should contain a column called `gRNA_group` indicating the "group" to which each gRNA belongs. Typically, targeting gRNAs are grouped according to the site that they target, and non-targeting gRNAs are grouped randomly into sets of size two or three.

```{r}
# in R
gRNA_modality
gRNA_modality |>
  get_feature_covariates() |>
  head()
```

-   The gene modality should be an integer-valued matrix of gene expressions.

```{r}
gene_modality
```

Next, the data frame containing the pairs to analyze should contain columns `gene_id` and `gRNA_group`. Additional columns are permitted but ignored.

```{r}
# in R
pairs_df <- readRDS(pair_fp)
head(pairs_df)
```

The gene IDs within the `gene_id` column should be a subset of the feature IDs of the gene modality; meanwhile, the gRNA groups within the `gRNA_group` column should be a subset of the entries of the `gRNA_group` column of the feature covariate matrix of the gRNA modality.

```{r}
# in R
all(pairs_df$gene_id %in% get_feature_ids(gene_modality)) # gene ID check
all(pairs_df$gRNA_group  %in% get_feature_covariates(gRNA_modality)$gRNA_group) # gRNA group check
```

## Output file

The location in which to store the results is specified by the output file path (`result_fp`).

## Additional arguments

The `sceptre` pipeline accepts several additional arguments, ordered here from most important to least important.

-   `formula`: an R formula (stored as a string) specifying the covariates to adjust for in the analysis. The variables in the string are assumed to be columns of the global cell covariate matrix. An example formula is as follows:

```{=html}
<!-- -->
```
    # in bash
    formula="~gene_p_mito+gene_batch+log(gene_n_nonzero)+log(gene_n_umis)+log(gRNA_n_nonzero)+log(gRNA_n_umis)"

The variables in this formula (`gene_p_mito`, `gene_batch`, `gene_n_nonzero`, `gene_n_umis`, `gRNA_n_nonzero`, and `gRNA_n_umis`) are columns of the global cell covariate matrix:

```{r}
# in R
crispr_experiment |> get_cell_covariates() |> head() 
```

The default behavior is to adjust for all (untransformed) variables stored within the global cell covariate matrix.

**Note**: The formula string should contain **no** white spaces (e.g., spaces, tabs, etc.).

-   `threshold`: the threshold to use to assign gRNAs to cells. For a given cell and gRNA, if the UMI count of the gRNA within the cell is equal to or greater than the threshold, then the gRNA is taken to be present in the cell. The default is 3.

-   `B`: the number of resamples to draw for the conditional randomization test. The default is 1000.

-   `side`: the sidedness of the test, one of "left", "right", or "both". The default is "both".

-   `n_pairs_to_sample`: the number of randomly-selected gene-gRNA group pairs on which to run `sceptre`. This parameter is for debugging purposes; often, it is useful to run the pipeline on (say) 25 randomly-selected pairs to ensure that the pipeline is set up correctly. The default is to run the pipeline on the entireset of pairs, i.e., to not subsample at all.

-   `gene_pod_size`, `gRNA_group_pod_size`, and `pair_pod_size`: parameters that control the extent to which the pipeline is parallelized. At a high level the pipeline works as follows: first, all genes are regressed onto the technical factors; next, all gRNA groups are regressed onto the technical factors; finally, all pairs of genes and gRNA groups (as specified in the pairs data frame) are tested for association. `gene_pod_size` (resp., `gRNA_group_pod_size`) is the number of genes (resp., gRNA groups) to regress onto the technical factors in a given Nextflow process. Meanwhile, `pair_pod_size` is the number of gene-gRNA group pairs to test for association in a given Nextflow process. The default values are 50, 50, and 100, respectively.

## Invoking the pipeline

An example invocation script is available in the `sceptre` Nextflow repository. Git clone the repository.

    # on command line
    git clone git@github.com:timothy-barry/sceptre-pipeline.git
    cd sceptre-pipeline

The bash script `example_launch.sh` contains an example invocation of the pipeline. We can launch this script via a call to `bash` on a laptop/desktop, `qsub` on a cluster running a Sun Gride Engine scheduler, or `sbatch` on a cluster running a SLURM scheduler, etc.

    # on command line
    bash example_launch.sh # laptop
    qsub example_launch.sh # slurm scheduler
